---
title: "Religion and Well-Being"
format: html

execute: 
  warning: false
  message: false
---

## Setup

Start an R Project for this activity within your private GitHub repo for this class. Put this file and the data file in the folder for the project. 

```{r}
## load any packages you need
library(tidyverse)
```

Read the data documentation, questionnaire, and project description files to better understand the data and the goals of this activity.

```{r}
religiosity <- read.csv("MARP_data.csv", stringsAsFactors = FALSE)
```

For each of the following two research questions, make a visual and statistical argument to support your answer.

## Research Question 1

**Do religious people self-report higher well-being?**

#### Identifying who is religious and who is not religious

```{r}
religiosity$is_religious = ifelse(religiosity$rel_3 != 1, 0, 1)

table(religiosity$is_religious)
```

#### Adding all the religious scores up to 9

```{r}
religiosity = religiosity %>% 
  mutate(rel_score = rowSums(across(rel_1:rel_9)))
```

#### Adding up 

```{r}
religiosity = religiosity %>%
  mutate(wb_score = rowSums(across(wb_gen_1:wb_soc_3)))
```


## Research Question 2

**Does the relation between religiosity and self-reported well-being depend on perceived cultural norms of religion?**

### EDA

```{r}

```



