---
title: "Religion and Well-Being"
format: html

execute: 
  warning: false
  message: false
---

## Setup

Start an R Project for this activity within your private GitHub repo for this class. Put this file and the data file in the folder for the project. 

```{r}
## load any packages you need
library(tidyverse)
```

Read the data documentation, questionnaire, and project description files to better understand the data and the goals of this activity.

```{r}
religiosity <- read.csv("MARP_data.csv", stringsAsFactors = FALSE)
```

For each of the following two research questions, make a visual and statistical argument to support your answer.

## Research Question 1

**Do religious people self-report higher well-being?**

#### Identifying who is religious and who is not religious

```{r}
religiosity$is_religious = ifelse(religiosity$rel_3 != 1, 0, 1)

table(religiosity$is_religious)
```

#### Adding all the religious scores up to 9 and getting the mean

```{r}
religiosity = religiosity %>% 
  mutate(rel_score = rowSums(across(rel_1:rel_9)), 
         rel_score_mean = rel_score / 9)
```

#### Adding up 

```{r}
religiosity = religiosity %>%
  mutate(wb_score = rowSums(across(wb_gen_1:wb_soc_3)), 
         wb_score_mean = wb_score / 18) 

# normalize from 1 to 5 to 0 to 1
religiosity = religiosity %>%
  mutate(wb_score_mean_norm = (wb_score_mean - 1) / 4)
```

#### Graphs

```{r}
religiosity$rel_3 <- case_when(
  religiosity$rel_3 == 0   ~ "Atheist",
  religiosity$rel_3 == 0.5 ~ "Agnostic",
  religiosity$rel_3 == 1   ~ "Religious"
)

ggplot(religiosity, aes(x = rel_score_mean, y = wb_score_mean_norm, color = rel_3)) + 
  geom_jitter() + 
  labs(title = "Mean Overall Religious Score by Mean Overall Well Being", 
       x = "Mean Religious Score", 
       y = "Mean Well Being Score", 
       color = "Religious Self Identification")
```

```{r}
cor(religiosity$rel_score_mean, religiosity$wb_overall_mean, use = "complete.obs")
cor(religiosity$rel_score_mean, religiosity$wb_phys_mean, use = "complete.obs")
cor(religiosity$rel_score_mean, religiosity$wb_psych_mean, use = "complete.obs")
```

There seems to be very little correlation between the mean religious score and the mean well being score. There is a slightly higher correlation between the mean overall psychological score, but overall it is not anywhere near enough to claim that there is a correlation. 
## Research Question 2

**Does the relation between religiosity and self-reported well-being depend on perceived cultural norms of religion?**

#### 

```{r}
religiosity = religiosity %>%
  mutate(c_norm_score = rowSums(across(cnorm_1:cnorm_2)), 
         c_norm_mean = c_norm_score / 2)

religiosity_long = religiosity %>% 
  pivot_longer(cols = c(rel_score_mean, c_norm_mean, wb_score_mean_norm), 
               names_to = "score_type", 
               values_to = "score")

ggplot(religiosity_long, aes(x = score, fill = score_type)) + 
  geom_density(alpha = 0.4) +
  scale_fill_manual(
    values = c(
      "c_norm_mean" = "skyblue",
      "rel_score_mean" = "#069494", 
      "wb_score_mean_norm" = "navy"
    )
  ) + labs(
    title = "Density Curve of Scores by Score Type", 
    x = "Normalized Score (from 0 to 1)"
  )
```

This plot does not show very much overlap in trends for all three of the scores.

```{r}
ggplot(religiosity, aes(x = rel_score_mean, y = wb_overall_mean, color = c_norm_mean)) + 
  geom_jitter() + 
  labs(title = "Mean Overall Religious Score by Mean Overall Well Being", 
       x = "Mean Religious Score", 
       y = "Mean Well Being Score", 
       color = "Religious Self Identification")
```

```{r}
cor(religiosity$rel_score_mean, religiosity$c_norm_mean, use = "complete.obs")
cor(religiosity$wb_score_mean_norm, religiosity$c_norm_mean, use = "complete.obs")
```

There seems to be a somewhat strong correlation between the mean religious score and the mean normal score. This would suggest that there is a correlation between the two, but when it comes to the relationship between religiosity and well being, it does not seem that there is much of a effect that cultural normal scores have. 

#### Mean of Means Score:

```{r}
mean(religiosity$rel_score_mean)
mean(religiosity$wb_score_mean_norm, na.rm = TRUE)
mean(religiosity$c_norm_mean)
```



