---
title: "Karen Training"
format: html

execute: 
  warning: false
  message: false
---

## Setup 

Start an R Project for this activity. Put this file in the folder for the project.

```{r}
library(tidyverse)
```

## Data

```{r}
england_wales_names <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-06-16/england_wales_names.csv')
ni_names <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-06-16/ni_names.csv')
scotland_names <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-06-16/scotland_names.csv')
```

```{r}
head(england_wales_names)
```

Note that the US Karen analysis normalizes the counts of each name/sex/year by the number of male/female babies per year. We will want to do the same here.

Here is some pseudocode to get you started.

- get number of babies per sex/year
- join that information with the original dataset
- make a new variable called `prop` that is the proportion of female/male babies in that year (rather than count)

## Normalize by number of female/male babies per year

```{r}
ni_names <- ni_names %>% 
  group_by(Sex, Year) %>% 
  summarise(full_count = sum(Number, na.rm = TRUE), .groups = "drop") %>% 
  left_join(ni_names, by = c("Year", "Sex")) %>% 
  mutate(prop = Number / full_count)
```

## Look at Karen across countries

To make a line per country, we need to put the three datasets together and have a column that specifies which country the data is from.

```{r}
england_wales_names$country = rep("England/Wales", nrow(england_wales_names))

## take same approach for Scotland
scotland_names$country = rep("scotland", nrow(scotland_names))

## take same approach for Northern Island
ni_names$country = rep("ni", nrow(ni_names))

## uncomment when you are ready
all <- bind_rows(england_wales_names, scotland_names, ni_names)
```

Now make a line plot over time per country of only the name Karen (female) using `ggplot`. Make sure to make informative axis labels and title. Does the pattern look like the Karen pattern in [the article](https://pudding.cool/2020/06/karen/)?

```{r}
only_karens = all %>%
  filter(Name == "Karen")

ggplot(only_karens, aes(x = Year, y = Number, fill = country, color = country)) + 
  geom_line(linewidth = 1) +
  labs(x = "Number of Babies Born",
       y = "Year",
       title = "Number of Babies Born named 'Karen' by Year")
```

## Add US data

It will be easier if we can just directly compare the datasets.

```{r}
#install.packages(babynames) ## run once if you don't have this package installed
library(babynames)

data("babynames")

## add a country column
babynames$country = "United States"

## uncomment the all line when ready
all <- all %>% select(Year, Sex, Name, Number, prop, country)
babynames = babynames %>%  rename(number = n) %>% select(year, sex, name, number, prop, country)
names(babynames) = tolower(names(babynames))
names(all) = tolower(names(all))

all <- bind_rows(all, babynames)
```

Now make a line plot over time per country of only the name Karen (female) again using `ggplot`. Make sure to make informative axis labels and title. Does the pattern look like the Karen pattern in the US?

```{r}


ggplot(all[all$name == "Karen" & all$country == "England/Wales", ], aes(x = year, y = number)) + geom_line()

ggplot(all[all$name == "Karen" & all$country == "scotland", ], aes(x = year, y = number)) + geom_line()

ggplot(all[all$name == "Karen" & all$country == "ni", ], aes(x = year, y = number)) + geom_line()

ggplot(all[all$name == "Karen" & all$country == "United States", ], aes(x = year, y = number)) + geom_line()

ggplot(all[all$name == "Karen", ], aes(x = year, y = number, fill = country, color = country)) + geom_line()

```

## Updating the US data

You'll notice the babynames dataset is not up to date. Download Resource 2 from the data source [here](https://catalog.data.gov/dataset/baby-names-from-social-security-card-applications-national-data). Put the folder in the directory for this project. Unzip it to find the data for 2018-2024. Look at the data's format outside of RStudio.

Delete all the files except those for 2018-2024. As you walk through the next two chunks of code, annotate each line with a comment explaining what it is doing and why.

```{r}
files_to_read <- list.files("names/")

files_to_read_path <- paste("names/",files_to_read, sep = "")

years <- 2018:2024

new_data <- lapply(files_to_read_path, read.table, header = F, sep = ",")
head(new_data[[1]])

for(i in 1:length(new_data)){
  names(new_data[[i]]) <- c("name", "sex", "number")
  new_data[[i]]$year = rep(years[i], nrow(new_data[[i]]))
  new_data[[i]] <- new_data[[i]] %>% select(year, sex, name, number)
}

new_data <- do.call("rbind", new_data)

```

## Add calculated columns

```{r}
## add prop column

## add country column

## cbind.data.frame current all and this new data
```


## Remake plot

Again make a line plot over time per country of only the name Karen (female) using `ggplot`. Make sure to make informative axis labels and title. Does the pattern look like the Karen pattern in the US?

```{r}

```


## Save your nice data

You will need this nice dataset for the next stage of this activity, so go ahead and save it, so you can directly read it in.

```{r}
## replace full_data with whatever you have called the dataset with
## 4 countries worth of data that is most up to date and has the correct
## columns

## uncomment when ready
#write.csv(full_data, "full-names-data-nice.csv", row.names = F)
```

